---
title: "Vehicle Fuel Efficiency — Multiple Linear Regression Analysis"
author: "Brian Hoang"
date: "2025-03-01"
output: html_document
---

## Overview

This analysis examines which vehicle characteristics are most strongly associated
with fuel efficiency (MPG) using multiple linear regression. The dataset contains
records on 392 vehicles produced between 1970 and 1982, sourced from the UCI MPG
dataset. The primary predictors examined are vehicle weight, horsepower, and engine
displacement.

---

## 1. Setup

```{r setup, message=FALSE, warning=FALSE}
# Load required libraries
library(tidyverse)  # data manipulation and piping
library(ggplot2)    # visualizations
```

---

## 2. Load and Inspect Data

```{r load-data}
# Load dataset directly from URL — no local file needed
auto <- read.csv("https://raw.githubusercontent.com/mwaskom/seaborn-data/master/mpg.csv")

# Preview the first few rows
head(auto)

# Check variable types and structure
str(auto)

# Check total number of observations
nrow(auto)

# Summary statistics for all variables
summary(auto)
```

---

## 3. Data Cleaning

A small number of observations have missing values for horsepower. These are removed
before modeling since horsepower is one of our primary predictors.

```{r clean-data}
# Drop rows with missing horsepower values
auto_clean <- auto |>
  drop_na(horsepower)

# Confirm final sample size (expect 392)
nrow(auto_clean)
```

---

## 4. Exploratory Data Analysis

Before fitting any models, scatterplots are used to visually assess the relationship
between MPG and each potential predictor.

```{r eda-plots, message=FALSE}
# MPG vs Weight
# Expect a negative relationship — heavier vehicles tend to use more fuel
ggplot(auto_clean, aes(x = weight, y = mpg)) +
  geom_point(color = "#2E86AB", alpha = 0.6, size = 2.5, shape = 16) +
  geom_smooth(method = "lm", se = TRUE, color = "#E84855", fill = "#E84855",
              alpha = 0.15, linewidth = 1.2) +
  labs(
    title    = "Fuel Efficiency vs Vehicle Weight",
    subtitle = "Heavier vehicles tend to consume more fuel",
    x        = "Weight (lbs)",
    y        = "Miles Per Gallon (MPG)"
  ) +
  theme_classic()

# MPG vs Horsepower
# Expect a negative relationship — more powerful engines consume more fuel
ggplot(auto_clean, aes(x = horsepower, y = mpg)) +
  geom_point(color = "#3BB273", alpha = 0.6, size = 2.5, shape = 16) +
  geom_smooth(method = "lm", se = TRUE, color = "#E84855", fill = "#E84855",
              alpha = 0.15, linewidth = 1.2) +
  labs(
    title    = "Fuel Efficiency vs Horsepower",
    subtitle = "Higher horsepower is associated with lower MPG",
    x        = "Horsepower",
    y        = "Miles Per Gallon (MPG)"
  ) +
  theme_classic()

# MPG vs Displacement
# Relationship looks weaker and less distinct compared to the other two predictors
ggplot(auto_clean, aes(x = displacement, y = mpg)) +
  geom_point(color = "#F4A261", alpha = 0.6, size = 2.5, shape = 16) +
  geom_smooth(method = "lm", se = TRUE, color = "#E84855", fill = "#E84855",
              alpha = 0.15, linewidth = 1.2) +
  labs(
    title    = "Fuel Efficiency vs Engine Displacement",
    subtitle = "Larger engines are linked to lower fuel efficiency",
    x        = "Displacement (cu. inches)",
    y        = "Miles Per Gallon (MPG)"
  ) +
  theme_classic()
```

---

## 5. Summary Statistics

```{r summary-stats}
# Summary statistics for the four key variables used in modeling
summary(auto_clean[, c("mpg", "weight", "horsepower", "displacement")])
```

---

## 6. Model Fitting

Two nested models are fit and compared. Because weight, horsepower, and displacement
tend to be correlated, including all three risks multicollinearity — where predictors
explain overlapping variation in MPG.

```{r full-model}
# Full model: weight + horsepower + displacement
model1 <- lm(mpg ~ weight + horsepower + displacement, data = auto_clean)

summary(model1)
```

```{r reduced-model}
# Reduced model: weight + horsepower only
model2 <- lm(mpg ~ weight + horsepower, data = auto_clean)

summary(model2)

# ANOVA partial F-test comparing the two nested models
# Tests whether adding displacement significantly improves model fit
anova(model2, model1)
```

The partial F-test produces p = 0.381, indicating that displacement does not
significantly improve the model once weight and horsepower are already included.
The reduced model (model2) is selected as the final model.

**Final model:** `mpg = 45.64 - 0.00579(weight) - 0.0473(horsepower)`

---

## 7. Model Diagnostics

Four standard diagnostic plots are used to check whether regression assumptions
are reasonably satisfied.

```{r diagnostics, message=FALSE}
library(patchwork)

# Build a dataframe of diagnostic values from the final model
df_diag <- data.frame(
  fitted    = fitted(model2),       # fitted (predicted) values
  residuals = residuals(model2),    # raw residuals
  std_resid = rstandard(model2),    # standardized residuals
  sqrt_std  = sqrt(abs(rstandard(model2))),  # for scale-location plot
  leverage  = hatvalues(model2)     # leverage values (hat matrix diagonal)
)

# Plot 1: Residuals vs Fitted
# Checks linearity — no clear pattern means assumption is reasonable
p1 <- ggplot(df_diag, aes(fitted, residuals)) +
  geom_point(alpha = 0.4, size = 1.5, color = "#2E86AB") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(se = FALSE, color = "red", linewidth = 0.8) +
  coord_cartesian(ylim = c(-15, 15)) +  # zoom in without removing data points
  labs(title = "Residuals vs Fitted", x = "Fitted values", y = "Residuals") +
  theme_classic()

# Plot 2: Q-Q Plot
# Checks normality of residuals — points close to the line indicate normality
p2 <- ggplot(df_diag, aes(sample = std_resid)) +
  stat_qq(alpha = 0.4, size = 1.5, color = "#3BB273") +
  stat_qq_line(color = "red", linetype = "dashed") +
  labs(title = "Q-Q Residuals", x = "Theoretical Quantiles",
       y = "Standardized Residuals") +
  theme_classic()

# Plot 3: Scale-Location
# Checks constant variance (homoscedasticity) — flat trend line is ideal
p3 <- ggplot(df_diag, aes(fitted, sqrt_std)) +
  geom_point(alpha = 0.4, size = 1.5, color = "#F4A261") +
  geom_smooth(se = FALSE, color = "red", linewidth = 0.8) +
  labs(title = "Scale-Location", x = "Fitted values",
       y = "√|Standardized Residuals|") +
  theme_classic()

# Plot 4: Residuals vs Leverage
# Identifies influential observations — high leverage + large residual is concerning
p4 <- ggplot(df_diag, aes(leverage, std_resid)) +
  geom_point(alpha = 0.4, size = 1.5, color = "#9B59B6") +
  geom_hline(yintercept = 0, color = "grey50", linetype = "dashed") +
  coord_cartesian(xlim = c(0, 0.06)) +
  labs(title = "Residuals vs Leverage", x = "Leverage",
       y = "Standardized Residuals") +
  theme_classic()

# Display all four plots in a 2x2 grid
(p1 + p2) / (p3 + p4)
```

---

## 8. Cook's Distance

Cook's distance measures how much influence a single observation has on the fitted
model. Values above 0.5 are generally considered influential.

```{r cooks-distance}
# Maximum Cook's distance across all observations
max(cooks.distance(model2))

# Well below the 0.5 threshold — no single observation is disproportionately influential
```

---

## Results Summary

- **Final model:** `mpg = 45.64 - 0.00579(weight) - 0.0473(horsepower)`
- **R² = 0.706** — the model explains approximately 70% of variation in MPG
- Both weight and horsepower are significant at p < 0.001
- Displacement was excluded based on the partial F-test (p = 0.381)
- Max Cook's distance = 0.037, well below the 0.5 threshold
- Mild heteroscedasticity detected but does not substantially affect interpretation
